---
title: "Section 9"
author: "Natalia Garbiras Díaz"
date: "March 24, 2017"
output: slidy_presentation
---

Today: MLE
===============================
- Stochastic and systematic components (example: logit)
- Deriving and plotting the likelihood function
- Finding the MLE and its SE
- Reporting MLE results

The setup
=====

>- We are interested in estimating parameters of conditional distributions (e.g., $P(y_i =1 \mid X)$).
>- Each random draw can be written as ($y_i$, $x_i$) $\rightarrow$ we want to estimate the conditional distribution of $y_i$ given $x_i$. 
>- In order to get the MLE, we need to first assume the underlying distribution of $y_i$ given $x_i$. That is why we have a *parametric* model with some finite number of unknown parameters that we need to estimate.
>- With MLE we are looking for the parameter that maximizes the joint density of ($y_1, y_2, \cdots, y_N$) evaluated in the data. 
>- + Notice we are not interested in $x_i$ so we do not specify a model for how $x_i$ is generated. Alternatively, we could think of $x_i$ as non-random (i.e., fixed features of a population).
>- Notice also that while $y_i$ is independent, it is not identically distributed (depends on $X\beta$). 


The DGP: Stochastic Component
===============================

PDF of a Bernoulli (binary) variable:

$$Y_i \sim Y_{Bern}(Y_i \mid \pi_i) = (\pi_i)^{y_i} (1-\pi_i)^{(1-y_i)}$$

where $y_i = \{0,1\}$. 

Simplifies to $\pi_i$ if $y_i=1$ and $(1-\pi_i)$ if $y_i=0$.

But what is $\pi_i$? For that we need a *link* function. 


The Logistic Link Function
===============================

Systematic component (CDF): 

$$\pi_i = F(x) = \Lambda(x) = \frac{1}{1+e^{-x}} = \frac{e^x}{1+e^x}$$

+ Note that $\Lambda(x) \in [0,1]$ for all $x$ 

We can program this:

```{r}
logistic <- function(x){ exp(x) / (1 + exp(x)) }

```

---

```{r}

curve(logistic(x), xlim=c(-8, 8), xlab="x", main="CDF of the logistic distribution",
      ylab="cumulative probability", col="blue", lwd=3)

```

---

$\Lambda$ is our link function: it filters the linear combination (i.e., our model with covariates) and returns 0-1 bounded probabilities.

So, with covariates, we have:

$$\Lambda = F(X\beta) = \frac{1}{1+e^{-X\beta}} = \frac{e^{X\beta}}{1+e^{X\beta}}$$

And: 

$$P(y_i=1 \mid \beta) = \pi_i= \frac{1}{1+e^{-X\beta}}$$

$$P(y_i=0 \mid \beta) = 1 - \pi_i= 1 - (\frac{1}{1+e^{-X\beta}})$$


---

A small dataset

```{r}

Y <- c(1, 0, 1, 0, 0, 1, 1)
x <- c(.5, .3, .45, .4, .25, .7, .9)

cbind(Y, x)

```

What is the joint probability distribution of these $n=7$ observations? 

---

$$P(y \mid \pi) = \prod_{i=1}^{n} \pi^{y_i} (1-\pi_i)^{1-y_i}$$

where $\pi_i = F(x_i\beta) = \frac{1}{1+e^{-x_i\beta}}$

And so the likelihood is $$L = P(y \mid X\beta) = \prod_{i=1}^{n} F(x_i\beta)^{y_i} (1-F(x_i\beta))^{1-y_i}$$

And the log-likelihood is $$ \ln L = \sum_{i=1}^{n} \big( y_i \times  \ln [F(x_i\beta)] + (1-y_i) \times \ln [1-F(x_i\beta)] \big)$$

----

We can program the likelihood in `R`:

```{r}

LL <- function(beta, x, y){
    
    odds <- beta * x
    
    rate <- 1 / (1 + exp(-odds))
    
    sum( y * log(rate) + (1 - y) * log(1 - rate) )
}

```

---

And we can plot the log-likelihood of our fake data:

```{r}

LL_example <- function(x){
    
    odds <- x * c(.5, .3, .4, .4, .25) # here x is a placeholder for our vector of betas
    
    rate <- 1 / (1 + exp(-odds))
    
    y <- c(1, 0, 1, 0, 0)
    
    sum(ifelse(y==1, log(rate), log(1 - rate)))
    
}

betas <- seq(-10, 10, by=.01)
ll_betas <- unlist(lapply(betas, FUN=LL_example)) # no intercept here

```

---

```{r, fig.height=4}

par(mfrow=c(1,2))
plot(betas, ll_betas, col="blue", type="l", lwd=3, ylab="log-likelihood")
plot(betas, ll_betas, col="blue", type="l", lwd=3, ylab="log-likelihood", xlim=c(-5, 5))

```

What's the value of $\beta$ tha maximizes the log-likelihood?


Getting the MLE (computation search).
====

+ For many MLE, we do not have an analytic or closed-form solution (under the assumptions of the OLS, we do have one for the linear model). 
+ But we know that the MLE is the maximizer of the log-likehood (and, thus, also maximizes the likelihood function). 

-----

```{r}

beta.start <- 0
out  <-  optim(beta.start, 
            fn=LL,
            x=x,y=Y,
            hessian=T,
            method="L-BFGS-B", # stands for "Limited-memory Broyden-Fletcher-Goldfarb-Shanno"
            control=list(fnscale=-1))
out

```

[Interested in what `optim` does?](https://www.ibm.com/developerworks/library/ba-optimR-john-nash/)

---

```{r}

mle <- out$par
mle

glm(Y~0+x, family=binomial(logit))

```

How to get a standard error for the MLE? 
====

(see Freedman Chapter 7, [Wooldridge Chapter 12](https://www.amazon.com/Econometric-Analysis-Cross-Section-Panel/dp/0262232588/ref=sr_1_4?s=books&ie=UTF8&qid=1490357174&sr=1-4&keywords=jeffrey+wooldridge), and [Greene Chapter 21](https://www.amazon.com/Econometric-Analysis-7th-William-Greene/dp/0131395386))

+ As previously mentioned, we often do not have an explicit expression for th MLE. 
+ We can rely on asymptotic theory (i.e., as $n \rightarrow \infty$) to get a sense of the statistical properties of our estimator (including it's asymptotic distribution). 
+ The asymptotic covariance matrix for the maximum likelihood estimator can be estimated by using the inverse of the Hessian evaluated at the maximum likelihood estimated (Greene 2011). 
+ The Hessian is simply the matrix of second derivates. In our case, it is just the second derivative of the log-likelihood function with respect to $\beta_{MLE}$. 
+ In our logit example, $$\mathbf{H}= \frac{\partial^2\ln L}{\partial\beta ~ \partial \beta'}= \sum_{i=1}^{n} \Lambda_i(1-\Lambda_i)X_iX'_i$$
+ Notice $\mathbf{H}$ is always negative definite so the log-likelihood is globally concave (i.e., we are finding a global maximum).
+ In the MLE, the Hessian is also called th **Fisher information matrix** noted with $I(\theta_o)$, where $\theta_o$ is the true parameter (see Freedman 2009 Chapter 7). 
+ The asymptiotic variance simplifies to (this also involves the Law of Large Numbers): $$\sqrt{n}(\hat{\theta}-\theta_o)\overset{d}{\to}N(0, I(\theta_o)^{-1})$$
+ Notice that the MLE is consistent and, therefore the above mean converges to 0. 
+ The Fisher information measures the average curvature of the log-likelihood around $\theta_o$
+ We can consistently estimate the Hessian with the sample analogue (i.e., $I({\hat{\theta}})^{-1}$). 

---

Now, let's get the variance-covariance matrix of our MLE in our observed data: 

```{r}
vcov <- -solve(out$hessian)
```

>+ What are the SEs? 

---

What are the SEs? 

```{r}
se<- sqrt(diag(vcov))
se
```
Notice we do not have an intercept nor covariates.

How should we present the results?
==========================

1. Simulate betas

```{r}

simbetas <- rnorm(100000, mle, sqrt(vcov)) 
par(mfrow=c(1,1))
plot(density(simbetas), col="slateblue", lwd=3)
```

---

2. Simulate predicted value

Let's say we get data for one more observation and we want to predict the probability of $y=1$

```{r}

new_obs <- .45 # our new data point

cov <- simbetas * new_obs
pred.p <- 1 / (1+exp(-cov))
mean(pred.p)


hist(pred.p, col="goldenrod", main="predicted probability")
abline(v=mean(pred.p), col="red", lwd=3)

quantile(pred.p, probs = c(0.25, 0.75))

```

---

3. Simulate expected values

```{r}

# for the binomial distribution, the expected value equals 
# the rate, so to get the expected value we can just take
# the expectation of the predicted probabilities (for other
# distributions this might not hold, and we might want to 
# do the simulation of the stochastic component as well)
# so here we could just do
mean(pred.p)

```

---

Simulate first differences

```{r}

new_obs
new_obs.better <- 0.65

cov <- simbetas * new_obs.better
pred.p.2 <- 1 / (1 + exp(-cov))

# to get the expected value, here we can again just take the 
# mean of the predicted probabilities
mean(pred.p.2)


#now lets get the first difference
firs.diff <- mean(pred.p.2) - mean(pred.p)
firs.diff 
```

An increase in the independent variable is associated with an increase in the outcome.

---

Predicted probabilities plot

How would we do the plot of predicted probabilities?

```{r}
# Let's take values of X from -1 to 1
x_sim <- seq(-1, 1, by=.01)

pred.prob <- matrix(NA, length(x_sim), 3)

for(i in 1:length(x_sim)){

    cov <- simbetas * x_sim[i]
    pred.p <- 1/(1+exp(-cov))
    
    pred.prob[i,1] <- mean(pred.p)
    
    # And we can simulate 90% CIs: 
    pred.prob[i,2] <- quantile(pred.p, probs=.05) # lower CI
    pred.prob[i,3] <- quantile(pred.p, probs=.95) # upper CI
}

head(pred.prob)
```

---

```{r}

plot(x_sim, pred.prob[,1], 
     col="slateblue", type="l", 
     lty=1, lwd=3, main="Predicted probabilities",
     xlab="x", ylab="predicted probability", ylim=c(min(pred.prob), max(pred.prob)))
lines(x_sim,pred.prob[,2], col="slateblue", type="l", lty=3, lwd=2 )
lines(x_sim,pred.prob[,3], col="slateblue", type="l", lty=2, lwd=2 )


```


Log odds
====

+ Odds ratio: probability of a "success" to probability of a "failure" $$\frac{p}{1- p} = e^{X\beta}$$
+ Log odds ratio: $X\beta$


